---
title: "HW1"
author: "Larry Cheng"
format: 
  html:
    embed-resources: true
---

[Link to repo](https://github.com/LarryC-24/Homework_1)

## Problem 1

### Part a
```{r}
library(tidyverse)
# read data and name columns
data <- read.csv("data.csv")
colnames(data) <- c("Season", "Cultivar", "Repetition", "Plant_Height", "Insert_First", "Stem_Num", "Legume_Num", "Grain_Num", "Grain_Pod", "Weight", "Grain_Yield")
```

### Part b
```{r}
# There are 320 rows (observations) in the dataset
print(nrow(data))
# There are 40 unique cultivars and 2 unique seasons in the dataset
print(length(unique(data$Cultivar)))
print(length(unique(data$Season)))
# Observations in each season
season_count <- data |> 
  group_by(Season) |> 
  count()
print(season_count)
```

### Part c
```{r}
# Correlation between PH and GY
print(cor(data$Plant_Height, data$Grain_Yield))
# Correlation between PH and GY for each season
season_cor <- data |> 
  group_by(Season) |> 
  summarize(correlation = cor(Plant_Height, Grain_Yield))
print(season_cor)
# Grain yield for tallest soybean
print(data[data$Plant_Height == max(data$Plant_Height),]$Grain_Yield)
# Percentage of observations with MHG greater than average
print(nrow(data[data$Weight>mean(data$Weight),])/nrow(data) * 100)
# Highest average grain yield cultivar
high_GY <- data |> 
  group_by(Cultivar) |> 
  summarize(ave_GY = mean(Grain_Yield)) |> 
  filter(ave_GY == max(ave_GY))
print(high_GY)
```

### Part d
```{r}
# Overall averages
overall_ave <- data |> 
  summarize(ave_Plant_Height = mean(Plant_Height), 
            ave_Insert_First = mean(Insert_First),
            ave_Stem_Num = mean(Stem_Num),
            ave_Legume_Num = mean(Legume_Num),
            ave_Grain_Num = mean(Grain_Num),
            ave_Grain_Pod = mean(Grain_Pod),
            ave_Weight = mean(Weight),
            ave_Grain_Yield = mean(Grain_Yield)) |> 
  mutate("Season" = "Overall", .before = 1)
# Season Averages
season_ave <- data |> 
  group_by(Season) |> 
  summarize(ave_Plant_Height = mean(Plant_Height), 
            ave_Insert_First = mean(Insert_First),
            ave_Stem_Num = mean(Stem_Num),
            ave_Legume_Num = mean(Legume_Num),
            ave_Grain_Num = mean(Grain_Num),
            ave_Grain_Pod = mean(Grain_Pod),
            ave_Weight = mean(Weight),
            ave_Grain_Yield = mean(Grain_Yield))
# Combine data frames
ave_table <- rbind(overall_ave, season_ave)
print(ave_table)
```

### Part e

$$
\begin{aligned}
H_0: \mu_{GY1} = \mu_{GY2} \\
H_A: \mu_{GY1} \neq \mu_{GY2}
\end{aligned}
$$

```{r}
# t-test with function
t.test(filter(data, Season == 1)$Grain_Yield, filter(data, Season == 2)$Grain_Yield)
# manual t-test
stats_GY <- data |> 
  group_by(Season) |> 
  summarize(mu = mean(Grain_Yield),
            sigma_sq = var(Grain_Yield),
            n = n())
t = (stats_GY$mu[1] - stats_GY$mu[2])/sqrt(stats_GY$sigma_sq[1]/stats_GY$n[1] + stats_GY$sigma_sq[2]/stats_GY$n[2])
p = 2*(1-pt(t,nrow(data)-2))
cat("t statistic:", t, "p-value:", p)
```
:::{.callout-tip icon="false"}
## Answer
Based on the results of the t-test, we fail to reject the null hypothesis that the grain yield between the two seasons does not differ at the significance level of 0.05.
:::

## Problem 2

### Part a
```{r}
# read data
data1 <- read.csv("adult.data", header = FALSE)
```

### Part b
```{r}
# clean names
colnames(data1) <- c("age", "workclass", "fnlwgt", "edu", "edu_num", "marital_status", "occupation", "relationship", "race", "sex", "cap_gain", "cap_loss", "hrs_per_week", "native_country", "income")
```

### Part c
```{r}
# replace missing values (?) with NA
data1[data1 == " ?"] <- NA
# clean white spaces
data1 <- data1 |> 
  mutate(across(where(is.character), ~ trimws(., which = "left")))
# count missing values in each column
print(colSums(is.na(data1)))
```

### Part d
```{r}
# sample size before filter
print(nrow(data1))
# sample size after filter
ten_hrs <- data1[data1$hrs_per_week >= 10,]
print(nrow(ten_hrs))
```

### Part e
```{r}
# removing rows with missing values
ten_hrs_clean <- na.omit(ten_hrs)
```

:::{.callout-tip icon="false"}
## Answer
It would be difficult to predict income using observations where data is missing, so they are removed.
:::

### Part f
```{r}
# filter observations
ten_hrs_filtered <- ten_hrs_clean |> 
  filter(edu != "Preschool", age >= 18, age <= 90)
```

### Part g
```{r}
print(nrow(ten_hrs_filtered))
```

## Question 3

### Part a
```{r}
nextHappy <- function(n) {
  if (round(n) != n | n < 0) {
    stop("please enter a positive integer")
  } # check for valid input
  n_chr <- as.character(n) # convert number to string to separate digits later
  n_digs <- as.numeric(strsplit(n_chr, split = "")[[1]]) # separates the digits in the character number, then converts each to a number
  sum_sq <- sum(n_digs**2) # squares each separated digit and sums them
  return(sum_sq)
}

# reproducing examples
nextHappy(19)
nextHappy(82)
```

### Part b
```{r}
happySequence <- function(n) {
  if (round(n) != n | n < 0) {
    stop("please enter a positive integer")
  } # check for valid input
  sequence <- c(n) # start vector containing sequence
  next_num <- nextHappy(n) # find next number in sequence
  while(!(next_num %in% sequence) & next_num != 1) {
    sequence <- c(sequence, next_num) # append next number to vector
    next_num <- nextHappy(next_num) # find next number in sequence
  } # end loop if n = 1 or if sequence loops to first number
  isHappy <- FALSE
  if (next_num == 1) {
    isHappy <- TRUE
    sequence <- c(sequence, next_num)
    return(list(sequence = sequence, isHappy = isHappy)) # return happy sequence and TRUE
  } # input is happy by checking if while loop ended because last number was 1.
  return(list(isHappy = isHappy)) # only return isHappy if it is FALSE
}

# reproduce examples
happySequence(19)
happySequence(4)
```

### Part c
```{r}
# count how many are happy
happy_count <- 0
for (num in seq(1:1000)) {
  if (happySequence(num)$isHappy) {
    happy_count <- happy_count + 1
  }
}
print(happy_count)
# find percentage
print(happy_count/1000)
# find longest sequence
longest_seq <- c()
for (num in seq(1:1000)) {
  if (happySequence(num)$isHappy) {
    if (length(happySequence(num)$sequence) > length(longest_seq)) {
      longest_seq <- happySequence(num)$sequence
    }
  }
}
print(longest_seq)
```

